When to ignore perfect information: how iterative learning can lead to worse decisions

### 2021-03-31 manuscript/appendix.Rmd
consolidate appendix

This appendix provides a fully reproducible walk-through of the implementation of the analyses considered in this paper.  Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.  However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].  As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.  While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous. Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].

@@

Here we provide annotated code necessary to completely reproduce all of the analysis presented in the main paper. This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse]. This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.

@@

@@

@@

Note that because the transition matrices are now altered after every action and observation, it is necessary to repeat the complete SDP calculation to determine the optimal action given the updated uncertainty.  While straight forward, this is what makes adaptive management for MDP problems particularly computationally intensive and thus poorly suited for handling a large space of candidate models. Nevertheless, our example of two models is most feasible, and even the larger example of 36 models (appendix B) is not unreasonable. Here I make use of the `mdplearning` R package, which provides a straight-forward implementation of this algorithm.

```{r}
adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
x0 <- which.min(abs(states - 6))

am1 <- adaptive_management(transitions[1:2], reward, discount, 
                           model_prior = c(0.99, 0.01), x0 = x0, 
                           Tmax = 50, true_transition = transitions[[3]], 
                            epsilon = 0.001, max_iter = 2000)
```

```{r}
am <- am1$df %>% mutate(belief = am1$posterior$V1,
                        stock = states[state],
                        quota = actions[action])
write_csv(am, "../data/am.csv")

@@

### Additional calculations for plots

@@

@@

>>>>

@@

Because the methods applied here: Stochastic Dynamic Programming, Iterative Forecasts assessed with Proper Scoring Rules, or Adaptive Management, are not themselves novel, the reader is referred to previous reviews and text books which can provide a thorough introduction to methods, in particular, @Marescot2013 and @Smith1981.  However, implementation of these approaches generally depends on numerical methods, for which even precise mathematical formulae or pseudo-code descriptions can be inadequate to specify unambiguously or to properly facilitate replication [@Barnes2010].  As such, this appendix includes computer code in the R language, which can also be found in the corresponding R-Markdown document in the paper's GitHub repository, <https://github.com/cboettig/bad-forecast-good-decision>.  While computer code can be more verbose and difficult to read than mathematical formulae presented here, it is also less ambiguous. Though R is already widely used among ecologists, the implementations concern mostly matrix algebra that can be translated to other languages, and the `MDPtoolbox` package used here is also implemented in Matlab [@Marescot2013; @MDPtoolbox].

Below, I provide annotated code necessary to completely reproduce all of the analysis presented in the main paper. This analysis is run in R [@R] uses `MDPtoolbox` [@MDPtoolbox] for solving Markov Decision Processes (MDP) using stochastic dynamic programming functionality, `expm` for matrix exponentials [@expm], and a few custom MDP functions provided by our package, `mdplearning` [@mdplearning].  We will also use `tidyverse`  packages for basic manipulation and plotting [@tidyverse]. This file is also available as an RMarkdown document [@rmarkdown] at  <https://github.com/cboettig/bad-forecast-good-decision>.  The code below generates the data tables required for each plot shown in the main paper.  Data are stored as `.csv` tables to be accessible to any relevant software program. Plotting commands used in the main paper are embedded in the paper's RMarkdown file at <https://github.com/cboettig/bad-forecast-good-decision>.

@@

# Model definitions

@@

We store the replicate predictions under each model at each timestep, along with the scores, in a `predictions.csv` table for plotting.

@@

Note that because the transition matrices are now altered after every action and observation, it is necessary to repeat the complete SDP calculation to determine the optimal action given the updated uncertainty.  While straight forward, this is what makes adaptive management for MDP problems particularly computationally intensive and thus poorly suited for handling a large space of candidate models. Nevertheless, our example of two models is most feasible, and even the larger example of 36 models is not unreasonable. Here I make use of the `mdplearning` R package, which provides a straight-forward implementation of this algorithm.

```{r}
adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
am1 <- adaptive_management(transitions[1:2], reward, discount, 
                           model_prior = c(0.99, 0.01), x0 = x0, 
                           Tmax = 50, true_transition = transitions[[3]], 
                            epsilon = 0.001, max_iter = 2000)
```

I also compare the adaptive management solution to an approach which still integrates uncertainty over both models, but treats the resulting policy as fixed for the duration of management, rather than adaptive _learning_ by updating probabilities.  This approach is sometimes referred to as uncertainty "planning," as it accounts for the same initial uncertainty over models but does not attempt to learn.   For consistency, I use the same model prior as in the learning case, which heavily favors model 1 to begin with.  Not surprisingly, the resulting policy is nearly identical to that under model 1 alone.

```{r}
# Compute policy using SDP over both models, given fixed prior beliefs
policy_planning <- memoise::memoise(mdp_compute_policy, cache = memoise::cache_filesystem("cache/"))
policy <- policy_planning(transitions[1:2],  
                          reward = reward, 
                          discount = discount,
                          model_prior = c(0.99, 0.01),
                          Tmax = 50, 
                          epsilon = 0.001, 
                          max_iter = 2000)
```

```{r}
# Simulate under true model using fixed planning policy
non_am <- mdp_planning(transitions[[3]],  
                       reward = reward, 
                       discount = discount,
                       model_prior = c(0.99, 0.01),
                       x0 = x0, 
                       a0 = 1,
                       policy = policy$policy,
                       Tmax = 50)  %>% 
  mutate(belief = 0.99, # belief in model 1 is fixed
         method = "planning")
```

We combine results of the two methods and store the resulting `data.frame` of stock size, quota, and belief in Model 1 over time as `am.csv` for later plotting.

```{r}
am <- am1$df %>%
  mutate(belief = am1$posterior$V1,
         method = "learning") %>%
  bind_rows(non_am) %>%
  mutate(stock = states[state],
         quota = actions[action]) %>%
  select(time, stock, quota, belief, method)

write_csv(am, "../data/am.csv")

@@

## Adaptive management over many models

We repeat the analysis above over a grid of $r$ and $K$ values.

```{r}
# compare over grid set of candidate r and K
rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
Ks <- c(0.5, 1, 1.5, 4, 10, 16)
sigma <- 0.1
n_models <- length(rs)*length(Ks)
pars <- expand.grid(rs, Ks)
names(pars) <- c("r", "K")
pars$sigma <- 0.1
closure <- function(r, K){
  function(x, h = 0){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
  }
}
many_models <-  lapply(1:n_models, function(i) closure(pars[i,"r"], pars[i,"K"]))

# transition matrices
many_transitions <- lapply(seq_along(many_models), function(i) 
  transition_matrices(many_models[[i]], states,  actions, sigma))
names(many_transitions) <- as.character(seq_along(many_models))
true_transition <- transition_matrices(f3, states, actions, sigma_g)

## Once again, initial belief favors "model 1"
model_prior <- c(rep(.01, n_models-1) / (n_models - 1), .99)

With the transition matrices defined for all 42 models, we perform adaptive management as before:

```{r}
adaptive_management <- memoise::memoise(mdp_learning, cache = memoise::cache_filesystem("cache/"))
am_many <- adaptive_management(many_transitions,  
                               reward = reward, 
                               discount = discount,
                               model_prior = model_prior,
                               x0 = x0, 
                               Tmax = 20, 
                               true_transition = true_transition, 
                               epsilon = 0.001, 
                               max_iter = 2000)

As before, we can compare the adaptive learning case to the policy which accounts for the initial uncertainty over all `r n_models` models, but does not seek to learn.

```{r}
policy_many <- policy_planning(many_transitions, reward = reward,  discount = discount,
                          model_prior = model_prior, Tmax = 20, 
                          epsilon = 0.001, max_iter = 2000)

non_am_many <- mdp_planning(true_transition,  
                       reward = reward, 
                       discount = discount,
                       model_prior = model_prior,
                       x0 = x0, 
                       a0 = 1,
                       policy = policy_many$policy,
                       Tmax = 20) %>% 
  mutate(belief = 0.99, # belief in model 1 is fixed
         method = "planning")

model1 <- dim(am_many$posterior)[2]

am_multi <- am_many$df %>% 
  mutate(belief = am_many$posterior[,model1],
         method = "learning") %>%
  bind_rows(non_am_many) %>%
  mutate(stock = states[state],
         quota = actions[action]) %>%
  select(time, belief, stock, quota, method)

write_csv(am_multi, "../data/am_multi.csv")
```

# Posteriors

```{r}
posteriors <- data.frame(r = rs, K = Ks, t(am_many$posterior)) %>%
  pivot_longer(starts_with("X"), 
               values_to = "probability",
               names_to = "time") %>% 
  mutate(time = as.integer(as.factor(time)))

i <- 20
r_marginals <- posteriors %>% group_by(time,r) %>% 
  summarise(prob = sum(probability), .groups = "drop") %>% 
  filter(time %in% i)

r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time)
```

## Marginal Posterior Probability evolution over K

```{r}
i = 20
K_marginals <- posteriors %>% group_by(time,K) %>% 
  summarise(prob = sum(probability), .groups = "drop") %>% 
  filter(time %in% i) 
K_marginals %>%
  ggplot(aes(K, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time) + scale_x_log10()

# Economic value tabulation

@@

# Growth models

@@

## Growth models in the many-models case:

```{r plot_models, fig.width=6, fig.height=4}
final_belief <- t(am_many$posterior)
final_belief <- final_belief[,dim(final_belief)[[2]]]
posteriors <- data.frame(model = as.character(1:length(many_models)), belief = final_belief)

model_set <-c(f3,  many_models)
names(model_set) <- c("true", 1:length(many_models))
d <- 
  map_dfc(model_set, function(f) f(states) - states) %>%
  mutate(state = states) %>% pivot_longer(names(model_set[-1]), "model") %>%
  left_join(posteriors)

d %>% 
  ggplot(aes(state, value, group=model, alpha = belief)) +
  #geom_hline(aes(yintercept = 0), lwd=1, lty=2) + 
  geom_line(lwd=1.5) + 
  geom_line(aes(state, true, alpha=1), col = "red", show.legend = FALSE) +
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x") 
```

### 2021-03-31 manuscript/appendix_B.Rmd
consolidate appendix

d %>% pivot_longer(names(model_set[-1]), "model") %>%
  mutate(model = as.numeric(model)) %>%
  ggplot(aes(state, value, col = model)) +
  geom_hline(aes(yintercept = 0), lwd=1) + 
  geom_point(alpha = .5) + 
  geom_line(aes(state, true), col = "red") +
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x") + viridis::scale_color_viridis()

>>>>

d %>% pivot_longer(names(model_set[-1]), "model") %>%
  mutate(model = as.numeric(model)) %>%
  ggplot(aes(state, value, col = model)) +
  geom_hline(aes(yintercept = 0), lwd=1) + 
  geom_point(alpha = .5) + 
  geom_line(aes(state, true), col = "red") +
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16)) +
  ylab(bquote(f(x) - x)) + xlab("x") + scale_color_grey()
```

### 2021-03-31 manuscript/manuscript.Rmd
consolidate appendix

abstract: |
  Model-based forecasts have been enabled by recent explosion of data and computational methods and spurred on by decision-makers appetite for forecasts in everything from elections to pandemic response. Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.

@@

```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicating the belief that Model 1 is correct (blue), lines indicating stock biomass over time, green triangles indicating harvest quotas set by adaptive management. Panel A: Each model is assigned an initial prior belief. To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.  Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2A. Panel B: given 36 candidate models over a broad range of parameters, adaptive management leads to models that overharvest even more, leading to even lower stock sizes and lower harvest quotas."}
am <- read_csv("../data/am.csv")
fig3a <- am %>%  filter(time <=20) %>%
  ggplot(aes(time, stock, col = belief)) + 
  geom_line(show.legend = FALSE) + 
  geom_point(show.legend = FALSE) +
  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
  ylab("stock size") + labs(subtitle="A. Two models") + ylim(c(0,11))

am_multi <- read_csv("../data/am_multi.csv")
fig3b <- am_multi %>%
  ggplot(aes(time, stock, col = belief)) + 
  geom_line() + 
  geom_point() +  
  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
  ylab("fish stock") + labs(subtitle="B. 36 models")+ ylim(c(0,11))

@@

<!-- Intuition, Figure 4 -->
The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. However, Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while Model 2 leads to over-harvesting.  The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.  Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.  While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.  This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.

>>>>

abstract: |
  Encouraged by decision makers appetite for future information on topics ranging from elections to pandemics, and enabled by the explosion of data and computational methods,  model based forecasts have garnered increasing influence on a breadth of decisions in modern society. Using a classic example from fisheries management, I demonstrate that selecting the model that produces the most accurate and precise forecast can lead to decidedly worse outcomes.

@@

```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicating the belief that Model 1 is correct (blue), lines indicating stock biomass over time, green triangles indicating harvest quotas set by adaptive management. Panel A: Each model is assigned an initial prior belief. To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.  Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2A. Panel B: given 42 candidate models over a broad range of parameters, adaptive management leads to models that overharvest even more, leading to even lower stock sizes and lower harvest quotas."}
am <- read_csv("../data/am.csv")
fig3a <- am %>%  filter(time <=20) %>% #
  ggplot(aes(time, stock, group = method)) + 
  geom_line(aes(lty=method), col = "gray40") + 
  geom_point(aes(col = belief), show.legend = FALSE, size=2) +
  geom_line(aes(y = quota, lty=method), col = "gray40") + 
  geom_point(aes(time, quota, col = belief), shape=2, size=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
  ylab("fish stock") + labs(subtitle="A.")

am_multi <- read_csv("../data/am_multi.csv")
fig3b <- am_multi %>%
   ggplot(aes(time, stock, group = method)) + 
  geom_line(aes(lty=method), col = "gray40") + 
  geom_point(aes(col = belief), show.legend = FALSE, size=2) +
  geom_line(aes(y = quota, lty=method), col = "gray40") + 
  geom_point(aes(time, quota, col = belief), shape=2, size=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
  ylab("fish stock") + labs(subtitle="B.")

@@

<!-- Intuition, Figure 4 -->
The reason for Model 1's seemingly contradictory ability to make good decisions but bad forecasts becomes obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4A], it is hardly surprising that no method exists which would not prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4B], while Model 2 leads to over-harvesting.  The explanation comes from noticing that the stock size corresponding to the maximum growth rate under Model 1 (the peak of the curve) falls at almost exactly the same stock size as that of the peak growth rate for Model 3.  Meanwhile, the peak of Model 2 occurs at a substantially lower stock size.  While the optimal control solution appears to depend only on a step-ahead forecast accuracy (indeed, the SDP solution method used here takes only step-ahead forecast probabilities as input, [@Marescot2013]), mathematical analysis showed long ago [@Reed1979] that the optimal solution for this problem depends only on keeping the biomass at the value responsible for the maximum growth rate.  This realization is quite general: for most decision problems, simple models will exist under which the optimal decision is the same as it would be for the true model, even when that simple model is wrong in most other ways.
